---
title: "Healtcare Appointment No-Show"
output:
  pdf_document: default
  word_document: default
  html_notebook: default
---

DESCRIPTION

A patient is considered to be a no-show when they fail to be present for a scheduled appointment. For any healthcare organization, no-shows lead to higher costs and underutilization of resources, which affects the quality of service healthcare organizations provide.

In order to solve this problem, the organizations need to be able to understand why no-show patients do so.


Task 1 : Explore the data to check for missing values or erroneous entries, comment on redundant features, and add additional ones if needed.

```{r}
# Import excel
library(readxl)
appointments <- read_excel("appointments.xls")
# check structure of the data
str(appointments)
```

```{r}
# column names
names(appointments)
```



```{r}
# convert date columns into character
appointments$AppointmentRegistration = as.POSIXct(appointments$AppointmentRegistration)
appointments$AppointmentDate = as.POSIXct(appointments$AppointmentDate)

# Convert columns to factors
appointments$Gender = as.factor(appointments$Gender)
appointments$Status = as.factor(appointments$Status)
appointments$Diabetes = as.factor(appointments$Diabetes)
appointments$Alcoholism = as.factor(appointments$Alcoholism)
appointments$HyperTension = as.factor(appointments$HyperTension)
appointments$Handicap = as.factor(appointments$Handicap)
appointments$Smokes = as.factor(appointments$Smokes)
appointments$Scholarship = as.factor(appointments$Scholarship)
appointments$Tuberculosis = as.factor(appointments$Tuberculosis)
appointments$Sms_Reminder = as.factor(appointments$Sms_Reminder)

summary(appointments)  # summary of the data
```

```{r}
# check for null values
sapply(appointments, function(x) sum(is.na(x)))
```

Task 2 : Create a new feature called HourOfTheDay, which will indicate the hour of the day at which the appointment was booked.

```{r}
# create 3 new features for AppointmentRegistration
appointments <- transform(appointments, Day_Reg = format(AppointmentRegistration, "%d"))
appointments <-transform(appointments, Month_Reg = format(AppointmentRegistration, "%m"))
appointments <- transform(appointments, Year_Reg = format(AppointmentRegistration, "%Y"))

appointments$Day_Reg = as.factor(appointments$Day_Reg)
appointments$Month_Reg = as.factor(appointments$Month_Reg)
appointments$Year_Reg = as.factor(appointments$Year_Reg)

summary(appointments)  # summary of the data
```

Task 3 : Identify and remove outliers from the age column and explain the reason behind the selected outlier treatment using an appropriate plot.

```{r}
# We see from the summary(appointments) that the minimum age is -1.
# So we check all the values less than 0.
appointments[appointments$Age < 0, ]
# It is only one and we will drop it.
appointments <-appointments[!(appointments$Age<0),]
summary(appointments$Age) # summary of the column Age
```


```{r}
# Plot
library(ggplot2)
ggplot(appointments, aes(x=Age)) + geom_histogram(bins=40)
```

Task 4 : Analyze the probability of showing up with respect to different features. Create a scatter plot and trend lines to analyze the relation between the probability of showing up with respect to age or hour of the day, and describe your findings.

```{r}
library(ggplot2)
library(dplyr)
appointments %>% group_by(Age) %>% 
  summarise(Show_Up_probability=sum(Status=="Show-Up")/n()) %>% 
ggplot(aes(x=Age, y=Show_Up_probability)) + geom_point() + geom_smooth(se=F)
```

We see that the older people show-up more often that younger ones.

Task 5 : Create a bar graph to depict the probability of showing up for diabetes, alcoholism, hypertension, TB, smokes, and scholarship.

```{r}
# check probability stats about these features
prop.table(table(appointments$Diabetes, appointments$Status))  
prop.table(table(appointments$Alcoholism, appointments$Status))
prop.table(table(appointments$HyperTension, appointments$Status))
prop.table(table(appointments$Tuberculosis, appointments$Status))
prop.table(table(appointments$Smokes, appointments$Status))
prop.table(table(appointments$Scholarship, appointments$Status))
```

```{r}
# Plots
library(gridExtra)
library(ggplot2)
g_Diabetes <- ggplot(appointments, aes(x=Diabetes, fill=Status)) + geom_bar(position="fill")
g_Alcoholism <- ggplot(appointments, aes(x=Alcoholism, fill=Status)) + geom_bar(position="fill")
g_Hypertension <- ggplot(appointments, aes(x=HyperTension, fill=Status)) + geom_bar(position="fill")
g_TB <- ggplot(appointments, aes(x=Tuberculosis, fill=Status)) + geom_bar(position="fill")
g_Smokes <- ggplot(appointments, aes(x=Smokes, fill=Status)) + geom_bar(position="fill")
g_Scholarship <- ggplot(appointments, aes(x=Scholarship, fill=Status)) + geom_bar(position="fill")

grid.arrange(g_Diabetes, g_Alcoholism, g_Hypertension, g_TB, g_Smokes,  g_Scholarship, ncol=2, top='Bar graphs for features')
```

Task 6 : Create separate bar graphs to show the probability of showing up with respect to male or female, day of the week, and SMS reminder columns and describe your findings.

```{r}
# Bar graph to show the probability of showing up with respect to male or female
ggplot(appointments, aes(x=Gender, fill=Status)) + geom_bar(position="fill")
```

It seems like the probability both for men and women to Show Up is similar.



```{r}
# Bar graph to show the probability of showing up with respect to SMS reminder
ggplot(appointments, aes(x=Sms_Reminder, fill=Status)) + geom_bar(position="fill")
```

It seems like Sms reminder is not important enough.


```{r}
ggplot(appointments, aes(x=Day_Reg, fill=Day_Reg )) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We see that the day with the  maximum probability to show-up in the month is the 24th.


Task 7 : Use different classification models to predict the show or no-show status based on the features that display the most variation in the probability of showing up.

Logistic Regression Model

```{r}
appointments_2 <- select(appointments, Age, Gender, Scholarship, HyperTension, Diabetes, Alcoholism, Handicap,Sms_Reminder, Status)
```

```{r}
appointments_2 <- mutate_at(appointments_2, vars(Status), as.factor)
```

```{r}
log_model <- glm(Status ~ . ,family = binomial(link = 'logit'),  data = appointments_2 )
```

```{r}
summary(log_model)
```

```{r}
# Logistic Regression with training and test data set
library(caTools)
set.seed(100)

split = sample.split(appointments_2$Status, SplitRatio = 0.70)

train = subset(appointments_2, split == TRUE)
test = subset(appointments_2, split == FALSE)
```

```{r}
logit_model <- glm(formula = Status ~ . ,  data = train, family =binomial(link = 'logit') )
summary(logit_model)
```

Task 8 : Evaluate the models and choose the best one for the data.

```{r}
fitted_p <- predict(logit_model,newdata=test,type='response')
head(fitted_p)
```

```{r}
pred_test <-  ifelse(fitted_p>0.5,1,0)
```

```{r}
tab <- table(predicted = pred_test, actual = test$Status)
tab
```

We see that there is nearly 80 % chance that the patient will show up.